PLAE web app enables powerful searching and multiple visualizations across one million unified single-cell ocular transcriptomes

### 2022-08-25 manuscript_00.Rmd
many text updates

---
title: 'PLAE enables quick and easy searching across one million ocular transcriptomes'
author:
  - Vinay Swamy:
      institute: bg
  - Zachary Batz:
      institute: nnrl
  - David McGaughey:
      institute:
        - bg
      correspondence: "yes"
      email: mcgaugheyd@mail.nih.gov
institute:
  - bg: Bioinformatics Group, Ophthalmic Genetics & Visual Function Branch, National Eye Institute, National Institutes of Health
  - nnrl: Neurobiology, Neurodegeneration & Repair Laboratory, National Eye Institute, National Institutes of Health
  - mgog: Medical Genetics and Ophthalmic Genomics Unit, National Eye Institute, National Institutes of Health

date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  word_document:
    reference_docx: word-styles-reference-01.docx
    fig_caption: yes
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
bibliography: zotero_plaeApp.bib
csl: investigative-ophthalmology-and-visual-science.csl
abstract: "PURPOSE: To create a high performance reactive web application to query gene expression across cell type, species, study, and other factors. METHODS: We updated the content and structure of the underlying data (single cell Eye in a Disk, scEiaD) and wrote the web application PLAE (https://plae.nei.nih.gov) to visualize and explore it. RESULTS: The new portal provides quick visualization of over a million individual cells from vertebrate eye-and body transcriptomes encomopassing four species, 60 cell types, six ocular tissues, and 23 body tissues across 37 publications. As a test of the value of this unified pan-eye dataset, we propose consistent cell type markers, replicate neurogenic and cone macula markers, and propose six new cone macula and peripheral markers . CONCLUSION: The PLAE v0.91 web app serves the pan-ocular and body dataset, scEiaD. This provides the eye community a powerful and quick means to test hypotheses on gene expression across a highly diverse community derived database."

@@

scEiaD_2020_v00 <- pool::dbPool(RSQLite::SQLite(), dbname =  glue('{data_dir_vGiga}/MOARTABLES__anthology_limmaFALSE___5000-transform-counts-universe-batch-scVIprojectionSO-8-0.1-50-5.sqlite'))
scEiaD_2020_v01 <- pool::dbPool(RSQLite::SQLite(), dbname =  glue('{data_dir_vPLAE}/MOARTABLES__anthology_limmaFALSE___4000-counts-universe-study_accession-scANVIprojection-15-5-0.1-50-20__pointRelease01.sqlite'))

@@

Single cell transcriptomics of the retina date to the early 2000s when researchers [@hagstromConePigmentGene2000b; @jabsEvidenceP2XP2X2000; @liLocalizationDopamineD1receptor2004; @lindqvistSingleCellRTPCR2002a; @reidRetinoschisinPhotoreceptorsecretedProtein2003]  used RT-PCR approaches to amplify select transcripts in small numbers of cells. Later, researchers used microarray platforms to analyze a broader array of transcripts[@cherryDevelopmentDiversificationRetinal2009; @gustincichGeneDiscoveryGenetically2004a; @kimIdentificationMolecularMarkers2008a; @roeschTranscriptomeRetinalMuller2008; @trimarchiIndividualRetinalProgenitor2008; @trimarchiMolecularHeterogeneityDeveloping2007b; @wahlinMethodAnalysisGene2004a]

The modern era of single cell transcriptomics is marked by the introduction of droplet-based technologies, which enabled quantitation of thousands of cells in a single experiment; in Macoksko et al. the adult mouse retina was profiled [@kleinDropletBarcodingSingleCell2015; @macoskoHighlyParallelGenomewide2015a]. The commercialization of the droplet approach has enabled broad access to this technology and the ocular community have responded by publishing several dozen ocular studies using single cell transcriptome technology since Macosko publication.

These publications's foci includes, but is not limited to, studying developmental transcriptomic dynamics, identifying genes to distinguish core cell types, proposing finer gradations of cell types within existing categories, and comparing ocular transcriptomes between organisms. Independent querying of these datasets requires a labrious series of steps. Briefly, one would have to identify the relevant publication, find the associated data deposit, download the cell - count matrix, find the cell label table (and optionally the t-SNE or UMAP coordinates), and then load the data into R or python to run the desired queries. If one desired to compare across publications, then one would also have to align the cell type labels and should re-create the count quantification with consistent bioinformatic tooling. It is desirable, perhaps even necessary, to use multiple datasets to assess gene expression to confirm whether a gene expression pattern is consistent and reproducible. Reliance on a single dataset can be problematic as there can be technical issues - for example the large consortium dataset GTEx has persistent contamination from highly expressed genes across tissues [@nieuwenhuisConsistentRNASequencing2020].

Web-based resources to query gene expression are powerful means to democratize access to large datasets as the computationally intensive steps above can be run once and then shared to anyone with a internet connected device. Single cell RNA-seq web resources that contain ocular tissues include the UCSC Cell Browser, Tabula Sapiens, PanglaoDB, and the Protein Atlas [@consortiumTabulaSapiensSingle2021; @franzenPanglaoDBWebServer2019; @karlssonSingleCellType2021; @speirUCSCCellBrowser2021]. Primary limitations to these resources include incomplete cell type labels and sample metadata that make ocular focused queries challenging. In contrast to these, the Spectacle resource is an ocular specific web app which has curated over a million cells from dozens of independent resources [@voigtSpectacleInteractiveResource2020]. However each data deposit is independently processed, which makes cross study queries difficult. We recently created a unified ocular data resource which placed dozens of independent single-cell RNA-seq ocular and non-ocular datasets in a single combined space [@swamyBuildingMegaSinglecell2021].

To broaden the usage of this powerful database, we restructured the data into a custom sqlite-based database and built a reactive web application around it. As many new studies were published since our last data build, we re-ran the scEiaD pipeline in March of 2022. In aggregate we have collated `r meta_filter$study_accession %>% unique() %>% length()` datasets across `r meta_filter$PMID %>% unique() %>% length()` publications, four species, and `r meta_filter$Tissue %>% unique() %>% length()` tissues. We hand-curated `r meta_filter$CellType %>% unique() %>% length()` cell type labels, totaling `r meta_filter %>% filter(!is.na(CellType)) %>% nrow()` cells and used those ground-truth labels as the basis for a machine learning-based algorithm which labels the remaining cells. Quality control metrics were tuned and applied to remove lower quality cells. In the end, the single cell Eye in a Disk (scEiaD) database contains `r meta_filter %>% nrow()` cells and 1,509,179,347 non-zero gene/cell expression values. We demonstrate how this unified resource can be used to identify genes that consistently distinguish cell types across studies and species, genes which have species-specific expression cell type patterning, demonstrate how our pre-computed differential testing can identify markers between similar progenitor cell types, and finally provide a R-based analysis document which end-to-end demonstrates how we use scEiaD to run a custom differential gene expression test to validate and propose new genes that distinguish macula from peripheral cones in humans.

@@

The scEiaD pipeline is largely the same as in Swamy et al. [@swamyBuildingMegaSinglecell2021]. Very briefly, we collected the raw fastq files and quantified with the kallisto / bustools system [@melstedBarcodeUMISet2019; @melstedModularEfficientPreprocessing2019]. The quantified counts are merged into one matrix that is used for downstream processing with scvitools create a batch corrected lower dimensional space. and then into Seurat for quality control, clustering, and 2D UMAP visualization. We use our previous xgboost-based machine learning approach to transfer cell type labels to unlabelled cells [@swamyBuildingMegaSinglecell2021].

The following alterations were made: First, the cutoff to retain a cell was raised from 200 unique genes quantified to 300. Second, we use the DecontX algorithm[@yangDecontaminationAmbientRNA2020] from the celda R package (version 1.9.2) to automatically remove ambient RNA contamination. Third, we altered the procedure for aligning gene names across species (see below for details). Fourth, we use the scANVI (scvitools based) batch correction method, which leverages the known cell type labels in the batch correction process [@xuProbabilisticHarmonizationAnnotation2021].

A custom cutoff for the minimum required number of unique genes identified per cell were used for the SRP362101 cornea dataset as an abnormally high number of cells were returned from our default value of 200. We set the cutoff for this studies to be 800 as this value resulted in approximately the same number of cells being returned as the authors reported in their papers.

@@

As we have four species in scEiaD (*Gallus gallus*, *Macaca fascicularis*, *Mus musculus*, and *Homo sapiens*) that we unify, we must identify shared genes. We found homologs/orthologs by pulling from BioMart database "Ensembl Genes 105" and using *Homo sapiens* as the reference. We then selected orthologous gene names from *Gallus gallus*, *Macaca fascicularis*, and *Mus Musculus*. This generates a table linked by Ensembl gene IDs and gene names. We then use a full join on non-duplicated Ensembl IDs between mouse and human. As we noticed a small number of gene names failed to be linked in this manner we ran another join on the remaining non-aligned genes, using gene name. This procedure gave us 17,769 shared human and mouse genes. To align the other species we again joined on Ensembl gene ID, removing genes where one chicken or macaque gene aligned to multiple human/mouse genes. In cases where multiple chicken or macaque genes aligned to one human/mouse gene, then we aggregated the counts by sum. In cases where there was no corresponding chicken or macaque gene, we filled in zeros. After these steps we had, as expected, 17,769 genes.

@@

Every study brought into scEiaD was hand curated to identify unique biological samples, published paper ID (where available), organ, tissue (e.g. Cornea), source (iPSC based or tissue), single cell platform (e.g. 10Xv2, DropSeq, etc.). We also, where possible, annotated retina region (e.g. macula), sex, and age or developmental stage. To import individual cell type labels (e.g. Rod, Muller Glia), we wrote custom code for each study that made this table available to link their cell type assignments back to the matching barcoded cell. We curated two types of cell type labels: "CellType" and "SubCellType" where the former is our normalized cell types taken from the published labels. We normalized names for CellType, for example changing "MG" to "Muller Glia" and also dropping more detailed cell type assignments like off and on bipolar cells for Bipolar Cells (as few studies went into this level). We did retain the original published (except for the broader name normalization) labels under SubCellType.

@@

The scater::aggregateAcrossCells function was used to aggregate gene counts across species, study accession, and [category] where category was either Cluster, CellType, or CellType (Predict) assignments for a cell [@mccarthyScaterPreprocessingQuality2017]. If there were fewer than 50 cells in organism - study accession - category combination, they were discarded for the differential test. Any genes with a sum of 0 counts after aggregation were removed. The aggregated matrix was imported into DESeq2 using the DESeqDataSetFromMatrix function with the design given as "~study_accession + [category]". Contrasts were extracted either with the target category (e.g. "Cone") against all remaining or in a pair wise manner ("Cone" vs "Rod") with the DESeq2 "results" function [@loveModeratedEstimationFold2014a]. 
<!-- Genes were considered to be significantly differentially expressed if they have a padj less than $1 x 10^-5 and a abs(log2FoldChange) > 2.  -->

@@

Our first version of the single cell Eye is a Disk (scEiaD) database contained `r meta_filter_v0$study_accession %>% unique() %>% length()` studies, three species, `r nrow(meta_filter_v0)` cells, and `r meta_filter_v0$CellType_predict %>% unique() %>% length()` curated cell types [@swamyBuildingMegaSinglecell2021]. We updated the scEiaD database for the PLAE v0.91 web app in six important ways. First, we added a new species, chicken (*Gallus gallus*) [@yamagataCellAtlasChick2021]. Second we added two ocular outflow tract datasets, one brain choroid plexus, and three cornea datasets to enhance coverage across the eye [@collinSingleCellAtlas2021; @gautamMultispeciesSinglecellTranscriptomic2021; @daniCellularSpatialMap2021; @ligockiMolecularCharacteristicsSpatial2021; @patelMolecularTaxonomyHuman2020; @vanzylCellAtlasAqueous2020]. Third, we added a human pan-body reference scRNA dataset to allow for non-ocular comparisons [@heSinglecellTranscriptomeProfiling2020]. Fourth, we used an *in silico* method to remove background gene contamination as we noticed persistent Rhodopsin expression in many non-photoreceptors cells (`r supFig_cap(name = 'decontX', display = 'cite')`). Fifth, we increased the cell retention cutoff for detected unique transcripts per cell from 200 to 300. Sixth, we carried over and harmonized several common cell type labels from the Tabula Muris project to more easily provide non eye cell type comparisons [@SinglecellTranscriptomics20].

@@

## scRNA web resources containing ocular data

@@

## PLAE differs from a complementary web app, Spectacle in several important ways

>>>>

---
title: 'PLAE web app enables powerful searching and multiple visualizations across one million unified single-cell ocular transcriptomes'
author:
  - Vinay Swamy:
      institute: bg
  - Zachary Batz:
      institute: nnrl
  - David McGaughey:
      institute:
        - bg
      correspondence: "yes"
      email: mcgaugheyd@mail.nih.gov
institute:
  - bg: Bioinformatics Group, Ophthalmic Genetics & Visual Function Branch, National Eye Institute, National Institutes of Health
  - nnrl: Neurobiology, Neurodegeneration & Repair Laboratory, National Eye Institute, National Institutes of Health
  - mgog: Medical Genetics and Ophthalmic Genomics Unit, National Eye Institute, National Institutes of Health

date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  word_document:
    reference_docx: word-styles-reference-01.docx
    fig_caption: yes
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
bibliography: zotero_plaeApp.bib
csl: investigative-ophthalmology-and-visual-science.csl
abstract: "PURPOSE: To create a high performance reactive web application to query single cell gene expression data across cell type, species, study, and other factors. METHODS: We updated the content and structure of the underlying data (single cell Eye in a Disk, scEiaD) and wrote the web application PLAE (https://plae.nei.nih.gov) to visualize and explore it. RESULTS: The new portal provides quick visualization of over a million individual cells from vertebrate eye-and body transcriptomes encomopassing four species, 60 cell types, six ocular tissues, and 23 body tissues across 37 publications. To demonstrate the value of this unified pan-eye dataset, we replicate known neurogenic and cone macula markers as well as propose six???? new cone human region markers. CONCLUSION: The PLAE web application provides the eye community a powerful and quick means to test hypotheses related to gene expression across a highly diverse, community-derived database."

@@

data_dir_vGiga <- '/Volumes/McGaughey_S/data/scEiaD/'
data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
scEiaD_2020_v00 <- pool::dbPool(RSQLite::SQLite(), dbname =  glue('{data_dir_vGiga}/MOARTABLES__anthology_limmaFALSE___5000-transform-counts-universe-batch-scVIprojectionSO-8-0.1-50-5.sqlite'))
scEiaD_2020_v01 <- pool::dbPool(RSQLite::SQLite(), dbname =  glue('{data_dir_vPLAE}/MOARTABLES__anthology_limmaFALSE___4000-counts-universe-study_accession-scANVIprojection-15-5-0.1-50-20__pointRelease01.sqlite'))

# Celltypes in celltype but not celltype_predict
dropped <- meta_filter$CellType[!meta_filter$CellType %in% meta_filter$CellType_predict]

@@

The retina is composed of six major cell populations: the photoreceptors, horizontal, bipolar, amacrine, ganglion, and non-neuronal cells [@masland_cell_2011]. These six populations can be further divided into dozens of cell types [@yan_cell_2020]. Furthermore, human disease of the eye can derive from damage to specific cell types. Glaucoma is characterized by damage to the retinal ganglion cells while cone and cone-rod dystrophies are characterized by photoreceptor degeneration. Efforts to untangle the transcriptomics of these diverse cell types with single cell transcriptomics date to the early 2000s when researchers [@hagstromConePigmentGene2000b; @jabsEvidenceP2XP2X2000; @liLocalizationDopamineD1receptor2004; @lindqvistSingleCellRTPCR2002a; @reidRetinoschisinPhotoreceptorsecretedProtein2003]  used RT-PCR approaches to amplify select transcripts in small numbers of cells. Later, researchers used microarray platforms to analyze a broader array of transcripts[@cherryDevelopmentDiversificationRetinal2009; @gustincichGeneDiscoveryGenetically2004a; @kimIdentificationMolecularMarkers2008a; @roeschTranscriptomeRetinalMuller2008; @trimarchiIndividualRetinalProgenitor2008; @trimarchiMolecularHeterogeneityDeveloping2007b; @wahlinMethodAnalysisGene2004a]

The modern era of single cell transcriptomics ha been marked by the introduction of droplet-based technologies, which enabled quantitation of thousands of cells in a single experiment. This approach was pioneered by Macoksko et al. and enabled the profiling of gene expression in 44,808 single cells in the adult mouse retina. [@kleinDropletBarcodingSingleCell2015; @macoskoHighlyParallelGenomewide2015a]. The commercialization of the droplet approach has enabled broad access to this technology and the ocular community have responded by publishing several dozen ocular studies using single cell transcriptome technology since Macosko publication.

These publications's foci includes, but are not limited to, studying developmental transcriptomic dynamics, identifying genes to distinguish core cell types, proposing finer gradations of cell types within existing categories, and comparing ocular transcriptomes between organisms. Independent querying of these datasets requires a laborious series of steps.

  Unfortunately manuscripts can only present a sliver of the underlying data and further exploration is laborious, technically challenging, and computationally expensive. Briefly, one would have to identify the relevant publication, find the associated data deposit, download the cell - count matrix, find the cell label table (and optionally the t-SNE or UMAP coordinates), and then load the data into R or python to run the desired queries. If one desired to compare across publications, then one would also have to align the cell type labels and should re-create the count quantification with consistent bioinformatic tooling. It is desirable, perhaps even necessary, to use multiple datasets to assess gene expression to confirm whether a gene expression pattern is consistent and reproducible. Reliance on a single dataset can be problematic as there can be technical issues - for example the large consortium dataset GTEx has persistent contamination from highly expressed genes across tissues [@nieuwenhuisConsistentRNASequencing2020].

Web-based resources to query gene expression can democratize access to large datasets as the computationally intensive steps above can be run once and then shared to anyone with a internet connected device. Single cell RNA-seq web resources that contain ocular tissues include the UCSC Cell Browser, Tabula Sapiens, PanglaoDB, and the Protein Atlas [@consortiumTabulaSapiensSingle2021; @franzenPanglaoDBWebServer2019; @karlssonSingleCellType2021; @speirUCSCCellBrowser2021]. Primary limitations to these resources include incomplete cell type labels and sample metadata that make ocular focused queries challenging. In contrast to these, the Spectacle resource is an ocular specific web app which has curated over a million cells from dozens of independent resources [@voigtSpectacleInteractiveResource2020]. However each data deposit is independently processed, which makes cross study queries impossible. We recently created a unified ocular data resource which placed dozens of independent single-cell RNA-seq ocular and non-ocular datasets in a single combined space [@swamyBuildingMegaSinglecell2021].

To broaden the usage of this powerful meta-atlas, we restructured the data into a custom sqlite-based database and built a reactive web application to access it. As many new studies were published since our last data build, we re-ran the scEiaD pipeline in March of 2022. In aggregate we have collated `r meta_filter$study_accession %>% unique() %>% length()` datasets across `r meta_filter$PMID %>% unique() %>% length()` publications, four species, and `r meta_filter$Tissue %>% unique() %>% length()` tissues. We hand-curated `r meta_filter$CellType_predict %>% unique() %>% length()` cell type labels, totaling `r meta_filter %>% filter(!is.na(CellType)) %>% nrow()` cells and used those ground-truth labels as the basis for a machine learning-based algorithm which labels the `r length(meta_filter$CellType_predict[!is.na(meta_filter$CellType_predict)]) - length(meta_filter$CellType[!is.na(meta_filter$CellType)])` more cells. Quality control metrics were tuned and applied to remove lower quality cells. In the end, the single cell Eye in a Disk (scEiaD) database contains `r meta_filter %>% nrow()` cells and 1,509,179,347 non-zero gene/cell expression values. We demonstrate how this unified resource can be used to identify genes that consistently distinguish cell types across studies and species, find genes which have species-specific expression cell type patterning, demonstrate how our pre-computed differential testing can identify markers between similar progenitor cell types, and finally provide a R-based analysis document which end-to-end demonstrates how we use scEiaD to run a custom differential gene expression test to validate and propose new genes that distinguish macula from peripheral cones in humans.

@@

The scEiaD pipeline is largely the same as in Swamy et al. [@swamyBuildingMegaSinglecell2021]. Very briefly, we collected the raw fastq files and quantified with the kallisto / bustools system [@melstedBarcodeUMISet2019; @melstedModularEfficientPreprocessing2019]. The quantified counts are merged into one matrix that is used for the downstream integration with scvitools create a batch corrected lower dimensional space. The raw counts and scANVI batch corrected lower dimensional space is sent into Seurat for quality control, clustering, and 2D UMAP visualization. We use our previously published xgboost-based machine learning approach to transfer cell type labels to unlabelled cells [@swamyBuildingMegaSinglecell2021].

The following alterations were made: First, the cutoff to retain a cell was raised from 200 unique genes quantified to 300. Second, we use the DecontX algorithm[@yangDecontaminationAmbientRNA2020] from the celda R package (version 1.9.2) to automatically remove ambient RNA contamination on a per-study basis. Third, we altered the procedure for aligning gene names across species (see below for details). Fourth, we use the scANVI (scvitools based) batch correction method, which leverages the known cell type labels in the batch correction process [@xuProbabilisticHarmonizationAnnotation2021].

A custom cutoff for the minimum required number of unique genes identified per cell were used for the SRP362101 cornea dataset as an abnormally high number of cells were returned from our default value of 300. We set the cutoff for this studies to be 800 as this value resulted in approximately the same number of cells being returned as the authors reported in their papers.

We found that while nearly all datasets we curated could be integrated as a whole, a small number had strange behavior in the two dimensional UMAP space and/or nearly all cells from a study were machine labelled as a single cell type. These studies were hand-removed from the resource.

@@

As we have four species in scEiaD (*Gallus gallus*, *Macaca fascicularis*, *Mus musculus*, and *Homo sapiens*) that we unify, we must identify shared genes. We found homologs/orthologs by pulling from the BioMart database "Ensembl Genes 105" and using *Homo sapiens* as the reference. We then selected orthologous gene names from *Gallus gallus*, *Macaca fascicularis*, and *Mus musculus*. This generates a table linked by Ensembl gene IDs and gene names. We then use a full join on non-duplicated Ensembl IDs between mouse and human. As we noticed a small number of gene names failed to be linked in this manner we ran another join on the remaining non-aligned genes, using gene name. This procedure gave us 17,769 shared human and mouse genes. To align the other species we again joined on Ensembl gene ID, removing genes where one chicken or macaque gene aligned to multiple human/mouse genes. In cases where multiple chicken or macaque genes aligned to one human/mouse gene, then we aggregated the counts by sum. In cases where there was no corresponding chicken or macaque gene, we filled in zeros. After these steps we had, as expected, 17,769 genes.

@@

Every study brought into scEiaD was hand curated to identify unique biological samples, published paper ID (where available), organ, tissue (e.g. Cornea), source (iPSC based or tissue), single cell platform (e.g. 10Xv2, DropSeq, etc.). We also, where possible, annotated retina region (e.g. macula), sex, and age or developmental stage. Samples with perturbations (genetic or treatment) were not included. To import individual cell type labels (e.g. Rod, Muller Glia), we wrote custom code for each study that made this table available to link their cell type assignments back to the matching barcoded cell. We curated two types of cell type labels: "CellType" and "SubCellType" where the former is our normalized cell types taken from the published labels. We normalized names for CellType, for example changing "MG" to "Muller Glia" and also dropping more detailed cell type assignments like off and on bipolar cells for Bipolar Cells (as few studies went into this level). We did retain the original published (except for the broader name normalization) labels under SubCellType.

@@

The aggregateAcrossCells function from the scuttle R package was used to aggregate gene counts across species, study accession, and category where category was either Cluster, CellType, or CellType (Predict) assignments for a cell [@mccarthyScaterPreprocessingQuality2017]. If there were fewer than 50 cells in organism - study accession - category combination, they were discarded for the differential test. Any genes with a sum of 0 counts after aggregation were removed. The aggregated matrix was imported into DESeq2 using the DESeqDataSetFromMatrix function with the design given as "~study_accession + category". Contrasts were extracted either with the target category (e.g. "Cone") against all remaining or in a pair wise manner ("Cone" vs "Rod") with the DESeq2 "results" function [@loveModeratedEstimationFold2014a]. 
<!-- Genes were considered to be significantly differentially expressed if they have a padj less than $1 x 10^-5 and a abs(log2FoldChange) > 2.  -->

@@

Our first version of the single cell Eye is a Disk (scEiaD) database contained `r meta_filter_v0$study_accession %>% unique() %>% length()` studies, three species, `r nrow(meta_filter_v0)` cells, and `r meta_filter_v0$CellType_predict %>% unique() %>% length()` curated cell types [@swamyBuildingMegaSinglecell2021]. We updated the scEiaD database for the PLAE v0.91 web app in six important ways. First, we added a new species, chicken (*Gallus gallus*) [@yamagataCellAtlasChick2021]. Second we added two ocular outflow tract datasets, one brain choroid plexus, and three cornea datasets to enhance coverage across the eye [@collinSingleCellAtlas2021; @gautamMultispeciesSinglecellTranscriptomic2021; @daniCellularSpatialMap2021; @ligockiMolecularCharacteristicsSpatial2021; @patelMolecularTaxonomyHuman2020; @vanzylCellAtlasAqueous2020]. Third, we added a human pan-body reference scRNA dataset to allow for non-ocular comparisons [@heSinglecellTranscriptomeProfiling2020]. Fourth, we used an *in silico* method to remove background gene contamination as we noticed persistent Rhodopsin expression in many non-photoreceptors cells (`r supFig_cap(name = 'decontX', display = 'cite')`). Fifth, we increased the cell retention cutoff for detected unique transcripts per cell from 200 to 300. Sixth, we carried over and harmonized several common cell type labels from the Tabula Muris project to more easily provide non eye cell type comparisons [@SinglecellTranscriptomics20]. As before, the pipeline's scVI's integration parameters were chosen rigorously by testing inegration performance across a wide range of latent dimensions and number of highly variable genes in our scPOP framework [@swamyBuildingMegaSinglecell2021]. Each study in the resource were hand assessed to determine whether they had any unusual properties in regards to machine learnt cell type proportions or position in the 2D UMAP space. The small number of outlier studies were removed from the resource.

@@

## PLAE resource contains several crucial features relative to other single cell resources

@@
